Elucidating the Transcriptome of Turkey Hemorrhagic Enteritis Virus

### 2024-01-06 manuscript_thev_transcriptome.Rmd
added E3 region plotand gels with SSC and STC

**Background:** Hemorrhagic enteritis (HE) is a disease affecting 6-12-week-old turkeys characterized by *immunosuppression (IS)* and bloody diarrhea. This disease is caused by *Turkey Hemorrhagic Enteritis Virus (THEV)* of which avirulent strains (THEV-A) that do not cause HE but retain the immunosuppressive ability have been isolated. The THEV-A Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. *Our objective is to understand the genetic basis by which VAS induces IS*. The *transcriptome of THEV* was studied to set the stage for further experimentation with specific viral genes that may mediate IS.\
**Methods:** After infecting a turkey B-cell line (MDTC-RP19) with the VAS vaccine strain, samples in triplicates were collected at 4-, 12-, 24-, and 72-hours post-infection. Total RNA was subsequently extracted, and poly-A-tailed mRNA sequencing done. After trimming the raw sequencing reads with the `FastQC`, reads were mapped to the THEV genome using `Hisat2` and transcripts assembled with `StringTie`. An in-house script was used to consolidate transcripts from all time-points, generating the final transcriptome. PCR, gel electrophoresis, and Sanger sequencing were used to validate all identified splice junctions.\
**Results and Conclusions:** A total of `r total_mapped` million reads mapped to THEV genome providing good coverage/depth, leaving no regions unmapped. All predicted genes in the genome were represented. In keeping with all adenoviruses, all transcripts were spliced with either with 5'- or 3'-multi exon UTRs hitherto unknown. ~~*Thirteen*~~ novel exons were identified which were validated by PCR and Sanger sequencing. The splicing patterns strongly suggest that there are ~~*three*~~ main promoters (E1, E3, and major late promoters) driving expression of most of the genes with ~~*two*~~ possible minor promoters driving single genes (ORF7 and ORF8). This RNA-sequencing experiment is the first study of THEV gene expression to date. In keeping with other Adenoviruses, almost all THEV genes are spliced, and several genes are expressed as one transcription unit under a single promoter. This insight into THEV's transcriptome may allow the engineering of the VAS to provide immune protection with less or no associated IS. \newpage

@@

Virulent strains (THEV-V) and avirulent strains (THEV-A) of THEV are serologically indistinguishable, infecting turkeys, chickens, and pheasants and the THEV-V cause different clinical diseases in these birds [@Harrach2008; @Beach2006]. In turkeys, the THEV-V cause hemorrhagic enteritis (HE), a debilitating acute disease affecting predominantly 6-12-week-old turkeys characterized by immunosuppression (IS), weight loss, intestinal lesions leading to bloody diarrhea, splenomegaly, and up to 80% mortality [@Beach2006; @Beach2009a; @Gross1967]. HE is the most economically significant disease caused by any strain of THEV [@Beach2006]. While the current vaccine strain (a THEV-A isolated from a pheasant, Virginia Avirulent Strain [VAS]) have proven effective at preventing HE in young turkey poults, it still retains the immunosuppressive ability. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated cohorts leading to substantial economic losses [@Beach2006; @Rautenschlein2000; @Larsen1985; @Dhama2017]. The induced IS also interferes with vaccination schemes for other infections of turkeys [@Beach2006; @Rautenschlein2000]. To eliminate this immunossupressive side-effect of the vaccine, a thorough investigation of the culprit viral factors (genes) mediating this phenomenon is essential. However, the transcriptome (splicing and gene expression patterns) of THEV has not been characterized, making the investigation of specific viral genes for possible roles in causing IS impractical. A well-characterized transcriptome of THEV is required to enable the next leap forward in THEV research - experimentation with specific viral genes that may mediate IS.

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs - specifically human AdVs - thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU), which generates five families of late mRNAs (L1-L5). An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs; all but pIX mRNA undergo at least one splicing event. The MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5). Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing of the genes encoded in the unit[@Zhao2014; @Donovan2020; @Guimet2016]. Almost all AdV mRNAs are generated by the excision of one or more introns and most of these introns are located in the 5' or 3' UTRs of pre-mRNA. Thus the viral introns scarcely interrupt the open reading frames (ORFs) [@Davison2003; @Zhao2014].

@@

**Overview of sequencing data and analysis pipeline outputs**\
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified an overall total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel, ~~and using 3' Rapid Amplification of cDNA Ends (3'RACE) and other methods, we further identified \#### unique splice variants~~. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (**Figure 3**). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under **accession no. XXXXXX**.

@@

**Early Region 1 (E1) transcripts**. This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (**Figure 5**). Most of the encoded proteins of the novel transcripts are distinct from the predicted ORFs, although they all have the potential to encode the Hyd protein as the 3'-most coding sequence (CDS) if secondary start codon usage is considered. The 5'-most CDS of TRXPT_1 is multi-exonic, producing a 17.9 kilodalton (kDa) protein of 160 residues [amino acids (aa)]. The CDS begins in the first exon, starting at position 211, spans the second exon, and terminates in the third exon at position 2312. From the 5'-most start codon (SSC), TRXPT_2 encodes the largest protein in this region - a 64.3 kDa, 580 aa protein with the same SSC as TRXPT_1 (position 211). This CDS spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 to 1964 (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of the Hyd STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. The N-terminus of TRXPT_2 CDS therefore, shares substantial protein sequence similarity with ORF1 but Hyd and TRXPT_2 are not in-frame; hence no protein sequence similarity. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. From our RNA-seq data, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2; however, given that studies in MAdVs show that E1 mRNAs share a common 5' and 3' positions, only differing from each other regarding the introns [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and the TSS (at position 54bp) just like the transcription termination site (TTS; at position 2325bp) are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential remains unaffected. Its 5'-most CDS, beginning at 1965 and sharing the same STC as TRXPT_1 and TRXPT_2, produces a 13.1 kDa, 115 residue protein. This CDS (ORF4) was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggests it is a bona fide ORF. The coding potential of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its coding potential is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS if the first SSC is skipped). However, if we assume that TRXPT_4 shares the same TSS as TRXPT_1, then the 5'-most CDS is a distinct multi-exonic 15.9 kDa, 143 aa protein with the same SSC as TRXPT_1 and TRXPT_2 but with a unique STC. All splice junctions of the transcripts in this region (except the junction for TRXPT_4) have been validated by cloning and Sanger sequencing of cDNA (**Figure 5b**; **supplementary PCR methods**). Finally, during our validation of TRXPT_2, ORF1 was present on the agarose gel and Sanger sequencing results as a bona fide transcript (**supplementary PCR methods**). This is corroborated by 3'-RACE experiment, which shows transcripts spanning the entire ORF1 and Hyd ORFs without any splicing with a poly-A tail immediately after the TTS of transcripts in this region. The 5'-most CDS of this transcript would encode ORF1. Given that the SSC of the predicted ORF1 is in-frame but downstream of TRXPT_2 SSC, it suggests that the predicted ORF1 CDS is incomplete; it shares the same SSC and TSS, and TTS as TRXPT_2, but has a unique STC.

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**. The E2 TU expressed on the reverse strand, is subdivided into E2A and E2B and encodes three classical AdV proteins: pTP and Ad-pol (E2B proteins), and DBP (E2A protein) [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters (E2-early and E2-late) were discovered [@Donovan2020], we discovered only a single promoter from which both E2A and E2B transcription is initiated. However, similar to MAdVs, E2A and E2B transcripts have distinct TTSs, and the E2B transcripts share the TTS of the IVa2 transcript of the IM region [@Donovan2020; @Zhao2014] (**Figure 6**). The E2A ORF, DBP is one of three THEV ORFs predicted to be spliced from two exons, with the CDS spanning both exons. The corresponding transcript (TRXPT_21) found in our data matches this predicted splicing pattern exactly but with a non-coding additional exon at the 5'-end (E2-5'UTR) at position 18,684-18,751 bp, making a three-exon transcript. The encoded protein (DBP; 380 residues, 43.3 kDa) remains identical. This transcript (TRXPT_21) was also corroborated in a 3'-RACE experiment. Additionally, from the 3'-RACE, a splice variant of TRXPT_21 which retains the second intron leading to a 2-exon transcript was found. This transcript (TRXPT_21B), albeit longer due to retaining the second intron and possessing a short 3' UTR, encodes a truncated isoform of DBP because the first SSC utilized by TRXPT_21, is followed shortly by STCs in the retained intron, and does not yield any viable product. The SSC 173 bp downstream of TRXPT_21's SSC yields a 346 residue, 39.3 kDa product, which is in-frame of DBP but wholly contained in the second exon starting at 18,013 bp and terminating just like DBP at 16,973 bp. TRXPT_21 and TRXPT_21B share a common TSS but TRXPT_21B as seen in our 3'-RACE data, extends 39 bp into an adenine-thymine (A-T) rich sequence where the polyadenylation occurs. This suggests that the bona fide E2A TTS is at 16,934 bp (**Figure 6**).

The E2B region transcripts share the TSS of E2A - begins with E2-5'UTR - but extend thousands of base pairs downstream to reach the TTS at 2334bp, which is immediately followed by an A-T rich sequence where polyadenylation probably occurs. Interestingly, the TTS of the E1 TU on the sense strand is in the immediate vicinity of this A-T rich sequence (position 2323-2339bp), and the A-T rich sequence is almost palindromic; hence it likely serves as the polyadenylation signal for both E1 and E2B transcripts. The E2B transcripts, TRXPT_6 and TRXPT_7 are almost identical except for an extra splice junction at the 3'-end of TRXPT_6, making TRXPT_6 a five-exon transcript and TRXPT_7, four (**Figure 6**). TRXPT_7 has the coding potential for IVa2 and both classical proteins (pTP and Ad-pol) encoded in this region, of which the pTP ORF is predicted to be spliced from two exons just like all other AdVs. The predicted splice junction is corroborated by our data; however, the transcript is marked longer than the predicted ORF. There are two novel exons, the third exon is significantly longer than predicted, and the last exon containing the bulk of the CDS is more than triple the predicted size, although the SSC and STC encoding pTP are unaffected. The first two exons are 5'-UTRs - the SSC here is immediately followed by STCs - hence, the 5'-most SSC (position 10,995bp) of the third exon which matches the predicted SSC is utilized. The encoded product is identical to the predicted pTP ORF (597 residues; 70.5 kDa). However, if we consider secondary SSC usage, the encoded product is identical to the predicted Ad-pol ORF (1112 residues; 129.2 kDa), with SSC at 6768bp and STC at 3430bp.

~~finish trxpt_7 and then discuss trxpt_6~~.

@@

The splice junction of ~~TRXPT_5 add trxpts from E2~~ were confirmed by cloning and Sanger sequencing of cDNA (**supplementary PCR methods**).

**Early Region 3 (E3) transcripts**.

@@

**Major Late Promoter Region (MLP) transcripts**. \newpage

@@

**Figure 6: E2 and IM region transcripts. a)** The splice map of the E1 and IM transcription units. Exons are depicted as boxes connected by introns (dotted lines). Red transcripts are generated from RNA-seq data and predicted ORFs are colored grey. Each transcript or ORF is labelled with its name to the right. The start codon (SSC) and stop codon (STC) of the 5'-most CDS of each transcript is indicated with the nucleotide position in brackets. The region of the virus is depicted at the bottom as a black line with labels of the nucleotide positions for reference **b)** The sequence reads covering the splice junctions are indicated with information about their validation status using cloning and Sanger sequencing.

>>>>

**Background:** Hemorrhagic enteritis (HE) is a disease affecting 6-12-week-old turkeys characterized by *immunosuppression (IS)* and bloody diarrhea. This disease is caused by *Turkey Hemorrhagic Enteritis Virus (THEV)* of which avirulent strains (THEV-A) that do not cause HE but retain the immunosuppressive ability have been isolated. The THEV-A Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. *Our objective is to understand the genetic basis by which VAS induces IS*. The *transcriptome of THEV* was studied to set the stage for further experimentation with specific viral genes that may mediate IS.\
**Methods:** After infecting a turkey B-cell line (MDTC-RP19) with the VAS vaccine strain, samples in triplicates were collected at 4-, 12-, 24-, and 72-hours post-infection. Total RNA was subsequently extracted, and poly-A-tailed mRNA sequencing done. After trimming the raw sequencing reads with the `FastQC`, reads were mapped to the THEV genome using `Hisat2` and transcripts assembled with `StringTie`. An in-house script was used to consolidate transcripts from all time-points, generating the final transcriptome. PCR, gel electrophoresis, and Sanger sequencing were used to validate all identified splice junctions.
\newpage

@@

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV are serologically indistinguishable, infecting turkeys, chickens, and pheasants and the THEV-V cause different clinical diseases in these birds [@Harrach2008; @Beach2006]. In turkeys, the THEV-V cause hemorrhagic enteritis (HE), a debilitating acute disease affecting predominantly 6-12-week-old turkeys characterized by immunosuppression (IS), weight loss, intestinal lesions leading to bloody diarrhea, splenomegaly, and up to 80% mortality [@Beach2006; @Beach2009a; @Gross1967]. HE is the most economically significant disease caused by any strain of THEV [@Beach2006]. While the current vaccine strain (a THEV-A isolated from a pheasant, Virginia Avirulent Strain [VAS]) have proven effective at preventing HE in young turkey poults, it still retains the immunosuppressive ability. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated cohorts leading to substantial economic losses [@Beach2006; @Rautenschlein2000; @Larsen1985; @Dhama2017]. The induced IS also interferes with vaccination schemes for other infections of turkeys [@Beach2006; @Rautenschlein2000]. To eliminate this immunossupressive side-effect of the vaccine, a thorough investigation of the culprit viral factors (genes) mediating this phenomenon is essential. However, the transcriptome (splicing and gene expression patterns) of THEV has not been characterized, making the investigation of specific viral genes for possible roles in causing IS impractical. A well-characterized transcriptome of THEV is required to enable the next leap forward in THEV research - experimentation with specific viral genes that may mediate IS.

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs - specifically human AdVs - thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU or major late promoter [MLP]), which generates five families of late mRNAs (L1-L5). An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. The MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5). Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing of the genes encoded in the unit[@Zhao2014; @Donovan2020; @Guimet2016]. During translation of AdV mRNA, recent studies strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020].

@@

**Overview of sequencing data and analysis pipeline outputs**\
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel, ~~and using 3' Rapid Amplification of cDNA Ends (3'RACE) and other methods, we further identified \#### unique splice variants~~. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (**Figure 3**). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under **accession no. XXXXXX**.

@@

**Early Region 1 (E1) transcripts**    
This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (**Figure 5**). Most of the encoded proteins of the novel transcripts are distinct from the predicted ORFs, although they all have the potential to encode the Hyd protein as the 3'-most coding sequence (CDS) if secondary start codon usage is considered. The 5'-most CDS of TRXPT_1 is multi-exonic, encoding a 17.9 kilodalton (kDa) protein of 160 residues [amino acids (aa)]. The CDS begins in the first exon, starting at position 211, spans the second exon, and terminates in the third exon at position 2312. From the 5'-most start codon (SSC), TRXPT_2 encodes the largest protein in this region - a 64.3 kDa, 580 aa protein with the same SSC as TRXPT_1 (position 211). This CDS spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 to 1964 (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of the Hyd STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. TRXPT_2 CDS therefore, shares substantial protein sequence similarity with ORF1 but not with Hyd, as the SSC of Hyd is not in-frame. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. From our RNA-seq data, all E1 transcripts share the same transcription termination site (TTS; at position 2325bp); however, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2 (at position 54bp). Given that studies in MAdVs show that E1 mRNAs share not only a common TTS but also the TSS, and only differ from each other regarding the introns [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and their actual TSS just like the TTS are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential (CP) remains unaffected. Its 5'-most CDS, beginning at 1965 and sharing the same STC as TRXPT_1 and TRXPT_2, produces a 13.1 kDa, 115 residue protein. This CDS (ORF4) was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggests it is a bona fide ORF. The CP of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its CP is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS if the first SSC is skipped). However, if we assume that TRXPT_4 shares the same TSS as TRXPT_1, then the 5'-most CDS is a distinct multi-exonic 15.9 kDa, 143 aa protein with the same SSC as TRXPT_1 and TRXPT_2 but with a unique STC. All splice junctions of the transcripts in this region (except the junction for TRXPT_4) have been validated by cloning and Sanger sequencing of cDNA (**Figure 5b**; **supplementary PCR methods**). Finally, during our validation of TRXPT_2, ORF1 was present on the agarose gel and Sanger sequencing results as a bona fide transcript (**supplementary PCR methods**). This is corroborated by 3'-RACE experiment, which shows transcripts spanning the entire ORF1 and Hyd ORFs without any splicing with a poly-A tail immediately after the TTS of transcripts in this region. The 5'-most CDS of this transcript would encode ORF1. However, the SSC of the predicted ORF1 is in-frame but downstream of TRXPT_2 SSC, suggesting that the predicted ORF1 CDS is a truncated ORF; it shares the same TSS, SSC, and TTS as TRXPT_2, but has a unique STC.

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**     
The E2 TU expressed on the anti-sense strand, is subdivided into E2A and E2B and encodes three classical AdV proteins essential for genome replication: pTP and Ad-pol (E2B proteins), and DBP (E2A protein) [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters (E2-early and E2-late) were discovered [@Donovan2020], we discovered only a single promoter from which both E2A and E2B transcription is initiated. However, similar to MAdVs, E2A and E2B transcripts have distinct TTSs, and the E2B transcripts share the TTS of the IVa2 transcript of the IM region [@Donovan2020; @Zhao2014] (**Figure 6**).

The E2A ORF, DBP is one of three THEV ORFs predicted to be spliced from two exons. The corresponding transcript (TRXPT_21) found in our data matches this predicted splicing pattern exactly but with a non-coding additional exon at the 5'-end (E2-5'UTR) at position 18,684-18,751 bp. Thus a three-exon transcript with an identical encoded protein (DBP; 380 residues, 43.3 kDa). This transcript (TRXPT_21) was also corroborated in a 3'-RACE experiment. Additionally, from the 3'-RACE, a splice variant of TRXPT_21 which retains the second intron leading to a 2-exon transcript was found. This transcript (TRXPT_21B), albeit longer due to retaining the second intron and possessing a short 3' UTR, encodes a truncated isoform of DBP because the first SSC utilized by TRXPT_21, is followed shortly by STCs in the retained intron, and does not yield any viable product. Utilizing the SSC 173 bp downstream of TRXPT_21's SSC yields a 346 residue, 39.3 kDa product, which is in-frame of DBP but entirely contained in the second exon. TRXPT_21 and TRXPT_21B share a common TSS but TRXPT_21B as seen in our 3'-RACE data, extends 39 bp into an adenine-thymine (A-T) rich sequence before the poly-A tail sequence occur, suggesting this position (16,934bp) as the bona fide E2A TTS (**Figure 6**).

The E2B region transcripts share the TSS of E2A - begins with E2-5'UTR - but extend thousands of base pairs downstream to reach the TTS at 2334bp in the IM region, which is immediately followed by an A-T rich sequence (position 2323-2339bp) where polyadenylation probably occurs. Interestingly, the TTS of the E1 region on the sense strand is in the immediate vicinity of this A-T rich sequence, and the A-T rich sequence is almost palindromic; hence it likely serves as the polyadenylation signal for both E1 and E2B/IM transcripts. The E2B transcripts, TRXPT_6 and TRXPT_7 are almost identical except for an extra splice junction at the 3'-end of TRXPT_6, making TRXPT_6 a five-exon transcript and TRXPT_7, four (**Figure 6**). TRXPT_7 has the CP for IVa2 and both classical proteins (pTP and Ad-pol) encoded in this region, of which the pTP ORF is predicted to be spliced from two exons just like in all other AdVs. The predicted splice junction of pTP is corroborated by our data; however, the full transcript is markedly longer than the predicted ORF: there are two novel 5' exons, the third exon (containing the SSC of pTP) is significantly longer than predicted, and the last exon containing the bulk of the CDS is more than triple the predicted size of pTP. The first two exons are 5'-UTRs because the SSC here is immediately followed by STCs; hence, the 5'-most SSC (position 10,995bp) of the third exon which matches the predicted SSC is utilized. The encoded product is identical to the predicted pTP ORF (597 residues; 70.5 kDa). If secondary SSC (secSSC) usage is considered, with SSC at 6768bp and STC at 3430bp, the encoded product is identical to the predicted Ad-pol ORF (1112 residues; 129.2 kDa).
~~finish trxpt_7 and then discuss trxpt_6~~.

@@

**Early Region 3 (E3) transcripts**.   
The E3 region is wholly contained in the MLTU and encodes proteins involved in modulating and evading the host immune defenses. In MAdVs, this region contains seven ORFs expressed from several transcripts which share the same TSS (from the E3 promoter) but have different TTSs [@Guimet2016; @Donovan2020; @Zhao2014]. However, some E3 transcripts use the MLP as their TSS. Due to sharing the same TSS, in MAdVs, secSSC usage is heavily relied on for gene expression in this region except for 12.5K and transcripts using the MLP as TSS, as utilizing only the first SSC cannot produce all the other transcripts in this TU [@Donovan2020].

In THEV, only one ORF (E3) was predicted in this region. However, we identified six transcripts here (`r paste0("TRXPT_", 22:27, collapse = ", ")`), all of which are novel (**Figure 7**). We identified two distinct TSSs - one similar to the classic MAdV E3 TSS (position 18,230bp) and the other about 500 bp downstream at 18,727bp. The E3 transcripts collectively have the CP for several THEV ORFs: 100K, 22K, 33K, pVIII, E3, Fiber (IV), and ORF7 belonging to the MLTU; however, some CDSs are nonidentical due to unpredicted splicing. For instance, 33K is one of the few THEV ORFs predicted to be spliced from two exons; however, we discovered it to be a four-exon ORF expressed from TRXPT_24. The first two exons were not predicted but the last two, match the predicted exons precisely and the CDS is in-frame. However, the first 20bp of the predicted 33K (including the SSC) is spliced as as part of the second intron of TRXPT_24. Thus, the bona fide 33K is a 19.8 kDa, 171 residue protein spanning four exons. TRXPT_24 can also encode the ORFs, pVIII and E3 if we consider downstream SSC usage. TRXPT_22 and TRXPT_23 share the TSS of TRXPT_24 (E3 promoter; 18,230bp) but have distinct TTSs. Considering the first SSCs, they both encode novel 73 residue proteins that is over 80% similar but differ at the C-termini. Considering downstream SSC usage, both transcripts can encode pVIII and E3 in that order, but TRXPT_23 being longer, has the CP for the Fiber ORF also. TRXPT_25 is the largest transcript in the TU. It also utilizes the classic E3 TSS but has distinct TTS. It is a two exon transcript, encoding a novel protein (t100K; 543 residues), which is a shorter isoform of the predicted 100K ORF. Considering secSSC usage on this transcript yields the predicted ORF, 22K. It also has the CP for pVIII and E3 in that order. Furthermore, during the validation of TRXPT_25's splice junction using primers that span the junction (18350-18717bp), we noticed a DNA band that corresponds to the full unspliced sequence (**supplementary PCR methods Table 1**). As TRXPT_25 only falls short of encoding the complete 100K protein due to its splice junction, this band (which we cloned and validated by Sanger sequencing) suggests that the predicted 100K is indeed expressed. This transcript (TRXPT_25B) although not seen in our RNA-seq data, likely shares the same TSS and TTS as TRXPT_25. Lastly, TRXPT_26 and TRXPT_27 share the same TSS, unique from the other transcripts in this region but have distinct TTSs. TRXPT_26 is a three-exon transcript but the first two are UTRs - only the terminal exon encodes proteins. It encodes pVIII as the 5'-most ORF and has the CP for E3 and Fiber in that order. TRXPT_27 on the other hand, is only a two-exon transcript but similarly, only the terminal exon contains the CDSs. It encodes Fiber as the 5'-most ORF and ORF7 downstream with secSSC usage. TRXPT_13, which seems to be an E3 transcript that uses the MLP as TSS is discussed under the MLTU transcripts.

@@

**Major Late Transcription Unit (MLTU) or MLP Region transcripts**. \newpage

@@

**Figure 6: E2 and IM region transcripts. a)** The splice map of the E2 and IM transcription units. Exons are depicted as boxes connected by introns (dotted lines). Red transcripts are generated from RNA-seq data and predicted ORFs are colored grey. Each transcript or ORF is labelled with its name to the right. The start codon (SSC) and stop codon (STC) of the 5'-most CDS of each transcript is indicated with the nucleotide position in brackets. The region of the virus is depicted at the bottom as a black line with labels of the nucleotide positions for reference **b)** The sequence reads covering the splice junctions are indicated with information about their validation status using cloning and Sanger sequencing.
